% Section 6: Implications
% Last updated from: v2_2025-01-12.md

\section{Implications for accountable AI systems}
\label{sec:implications}

\foam{} reframes explanation as a structured record designed to support contestation rather than a post-hoc narrative. Instead of producing a single rationale, the system outputs (i) an auditable argument structure (claims, warrants, rebuttals), (ii) explicit perspective configurations, and (iii) sentence-level provenance linking each substantive claim to a checkable source span. This shifts accountability from ``did the explanation sound plausible?'' to ``which premises and evidence does the output depend on, and where can a challenge be lodged?''

Operationally, \foam{} supports contestation at three levels~\cite{alfrink2023contestable}: (1) \textbf{evidence disputes} (a cited sentence does not support the tagged claim; missing counterevidence), (2) \textbf{inferential disputes} (the warrant connecting evidence to conclusion is invalid or incomplete), and (3) \textbf{normative disputes} (the perspective/value configuration is illegitimate or incomplete for the context). Because these objects are explicit, a reviewer can localize disagreement to specific nodes and request revision without reopening the entire output as free-form prose.

\textbf{Template trees as contestability scaffolds.} Each typed syllogism component---uniqueness, link, impact---corresponds to a discrete contestation target with evidentiary bindings. The traversal log records \textit{what} claims were made and \textit{why} the structure took this form, enabling evidence, structural, and normative challenges at the level of epistemic assumptions, not only conclusions.

Institutionally, the resulting artifact functions as an auditable dossier that can support downstream review within existing governance workflows (internal review, incident response, assurance audits); dispute resolution itself requires institutional process beyond what the technical system provides. The technical contribution is not replacing due process, but supplying the structured, traceable materials that make procedural review feasible at scale.
